---
title: "Assignment 1"
author: "Replace with name"
date: "Replace with date"
output: word_document
---

```{r eval=FALSE, include=FALSE}
This template includes information about the requiered chunks and the prefered format of text, diagrams and tables. All chunks with information only, are set to not produce any text in the output file.

Always start with saving this template to your designated fplder with a new name.

You must always include a chunk which is loading all packages that you use during the kniting session.

You must also always include a chunk with scripts that import all the data that is needed for the kniting session. This can either be one single chunk with many separate scripts or several chunks with one script each.
In this assignment you will need one dataset named Chlamydia. Be sure to include the last version of the datset including all variables.

```

 
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(RcmdrMisc) #this chunk start all functions of RCommander
#library(multcomp) #needed for posthoc test after ANOVA. Remove the # to activate
#library()if you have used other packages you may need to include them here
```


```{r, message=FALSE, warning=FALSE, include=FALSE}
#include a chunk that opens your data in R. If you import the data in RCommander you can copy-paste the chunk that is produced in Rcommander directly into this RMarkdown file 
```


## Introduction

In all scientific work we need good descriptions of real patterns to be able to understand the processes behind them. Therefore, the aim of this assignment is to practice the "data and researcher get together process", and also to give you a first glimpse of the power of **R**.  

This assignment address the following learnig outcomes

* knowledge of basic statistical concepts: independent observations, maximum likelihood, linear model, analysis of variance/deviance, error distributions.

* an understanding of the importance of independent observations in sampling data for analysis. 


## Method

In this assignment we first look at some basic **R** functions and different options for working with **R** (Golisher 2008). We make a dataset by downloading Chlamydia incidence data from the Public Health Agency of Sweden (2021). This data is used in RStudio and Excel to produce diagrams showing the develpment of Chlamydia in Sweden from 2011 - 2021

## Result

```{r eval=FALSE, include=FALSE}
Include the important result from your analyses.
All diagrams need to have a Figure legend. The figure legend kan be either written below the diagram chunk, or be included in the first row in the chunk with the command fig.cap="".
You can also controll the size of the diagrams with fig.height= and fig.width=

Example {r, fig.height=7, fig.width=7, fig.cap="Fig.1 Everything you write in between the quotation marks will be added as a figrue caption under the figure"}

Have a short informative running text explaining the results.

Add scripts in R chunks. Additional chunks can be included with the "insert"" button
```





## Discussion

In this report there will be no discussion. Instead all result is commented in the result section 

## References

Golisher D 2008. R for absolute beginners https://duncanjg.files.wordpress.com/2008/11/rcoursedraft1.pdf
Public Health Agency of Sweden (2021). Chlamydia data. https://www.folkhalsomyndigheten.se/folkhalsorapportering-statistik/statistik-a-o/sjukdomsstatistik/klamydiainfektion/?t=county 

